Choosing the Choice: Reflections on Modelling Decisions and Behaviour in Demographic Agent-Based Models

### 2017-03-23 article.tex
updated article to submitted version

\title{Choosing the Choice: Reflections on Modelling Decisions and Behaviour in Demographic Agent-Based Models}
\author{Jonathan Gray \and Jason Hilton \and Jakub Bijak}

@@

We suggest that inclusion and operationalisation of context are best tackled through an iterative process of development, and evaluation against empirical data. This approach places considerable demands on the modeller in terms of developing the model, but also in exercising their judgement as to what should be included, and identifying where a component should be iterated out. In a sense, this is a similar process to the stepwise approaches commonly used in constructing regression models. Here, however, it is considerably less amenable to automation, both because augmenting a model is time consuming, and because evaluating complex models poses considerable challenges. The challenge of evaluation can be somewhat ameliorated by the use of techniques like sensitivity analysis (see \citet{Thiele2014} for a review of several techniques, and \citet{Oakley2004} for an alternative approach), which can support the modeller in identifying how the individual parameters of a model contribute to overall outputs. This too has limitations, because as we discuss later in relation to heterogeneity in agents (section \ref{heterogeneity-in-decision-strategies}), context can take the form of both process, and parameter.
As a result, the onus remains on the modeller to use their best judgement.

@@

As we have intimated, we believe that introducing such additional complications to a model necessitates an iterative approach. We also believe that the implementation model of decision making should be treated as a feature to iterate on, and that different models should be contrasted against one another. This suggests that modularity should be a key consideration in model design, to support the modeller in evaluating the relative merits of different decision models, given the purpose of the overall model.

@@

That said, as demographers, we are generally concerned with decisions which affect the life-course. These necessarily have consequences which extend far beyond the present. For example, the decision to have a child may affect not only a parent's immediate satisfaction with life, labour market participation, income streams, consumption and so forth, but will continue to do so for the rest of their life. Future `opportunity costs' must also be borne in mind: taking such a decision now may widen or narrow future options in various other areas of life. Similar considerations are salient for migration and partnership decisions, and indeed for many other demographic and social phenomena, for which the questions of time and sequencing of events tend to be too often ignored (see \citet{Abbott2001} for an excellent overview).

@@

Here, \(U^t\), denoting present utility at time \(t\), is a function of the stream of individual rewards \(u_t\) at each time point, weighted  according to \(\delta^{\tau}\), with \(\delta\) being the discounting factor. This form has nice properties in that it is time-consistent; agents who prefer one option over another today will do so until the reward has been realised. As an example, consider a utility maximising migrant who optimises their lifetime utility by either settling in the country to which they have migrated, or saving for a time and then returning home in order to spend such savings. If such a migrant discounts future rewards exponentially, then they will make a choice between these alternatives based on the rate of discount and the value of their savings at home relative to consumption in the host country. Crucially, if the situation does not change, such a decision will remain optimal throughout the agents lifespan.

@@

%\citet{Sozou1998} provides an interesting interpretation of this behaviour which allows it to be recast as rational. If a reward can be considered as uncertain and subject to constant hazard of being `lost', then we obtain an exponential function, as is well known. If the underlying hazard is also unknown, and an agent continuously updates an estimate of it according to Bayes' rule, we obtain hyperbolic discounting. As time passes, more information about the underlying hazard rate is gained through observing that the reward is still available, and thus both the estimate and the variance of hazard rate diminishes, rapidly increasing the present value of the pay-off.

Such behaviour can be approximated by the use of quasi-hyperbolic or present-weighted exponential discounting, which, as the names suggests, accounts for time-inconsistency by including a weight on present rewards in the function, either fixed or variable \citep{Benhabib2010}.

@@

At a more fundamental level, \citet{Boyer2008} describes findings that suggest how we think about future rewards is intrinsically linked to memory, in that we predict how we would respond to future benefits by considering our past and our emotional response to similar situations. Furthermore, he argues that our ability to exercise self-restraint is a function of this ability to conduct ``mental time travel'': we stave off temptation by remembering how bad we felt last time we yielded to it.

To give a demographic example of how these models of time-preference could be important, the work of  \cite{Wrede2011} is instructive. He investigated the effect of hyperbolic discounting on fertility in a simple three period model where mothers could have both immediate rewards from motherhood, and delayed rewards in the form of being looked after in old age. He found that the introduction of hyperbolic discounting reduces numbers of births if this second, investment, motive is dominant, but that the ability to offset the immediate costs of motherhood through the financial tools might mitigate these effects. The idea of pre-commitment is also discussed; in this context, an individual might limit their future freedom of action through sterilisation. Agent based models are flexible enough to embed such time-preference models in their wider social context (discussed in \ref{sec:context_matters}).

@@

%For instance, it could be argued that practising pre-commitment is not very different from having exponential-discounting preferences in the first place\footnote{We are thankful to Frans Willekens for this comment.}. We believe that conceptually it is different, in that it relies on forecasting future behaviour, both in terms of the likelihood of changing one's mind in the first place, and secondly with the potential for errors due to incorrectly inferring future responses to the self-enforced costs. Furthermore, it restricts actions

@@

As well as differences between agents, theories of frame and script selection also suggests difference within single agents over time; difference frames and scripts may be selected in response to different contexts, developing experiences, or influences from peers. This leads to discussion of another form of difference in decision making which may be important to demographers, that of differences over the life course. Migrants who were once intending to save and return home may change their mind and settle permanently in response to a new interpretation of their identity and circumstances.

@@

All of these problems are far from trivial, particularly as there is little consensus about which models of decision are best in which contexts. How then should we proceed with modelling when the fundamental questions about how individuals make decisions, how they treat time, and how they differ between themselves in these factors are still very uncertain? One way forward is to use the same simulation set-up to analyse multiple models of behaviour \citep{Rossiter2014}. This multi-model approach allows the modeller to attempt to distinguish between more and less plausible models of behaviour in the particular research context. Conditional on the simulation set-up being appropriate, more plausible models may be better able to reproduce multiple empirical patterns at varying scales \citep{Werker2004, Bianchi2008}.

This commitment to analysing multiple models of behaviour within a single simulation project is facilitated by a modular approach to simulation design, so that different models of behaviour can be swapped in and out without changing the underlying conditions of the simulation, as can be seen in \citet{Gray2016}. Such an approach reflects standard object-oriented design principles, but is not always reflected in an academic context (see also \citet{Rossiter2015} for discussion of the application of software engineering principles to social simulation). Modularity also allows for the possibility of easy extension by other researchers with access to the code, allowing for further behavioural models to be examined within the same simulation framework.

@@

Second, the existing methods of the statistical design of experiments (e.g. \citeauthor{Chaloner} \citeyear{Chaloner}) would allow for adopting a systematic approach to model design and data collection. In this way, following the suggestions of \citet{Courgeau}, the simulations would be built iteratively, by applying the experimental design principles to identifying the elements of the model which require additional empirical information, and then by collecting this information.  In this way, the important aspects of the decision process discussed above – context, time and heterogeneity – can be explicitly built into the model, through its specification or parameters. Then, they can be learnt about through repeated collection of information – including experimental one. The modularity of the model would help this process, as it would allow for fine-tuning the different aspects of the model independently, before they can be assembled together. An outline of the process is illustrated in Figure \ref{fig:approach}.

>>>>

\title{Choosing the Choice: Reflections on Modelling Decisions and Behaviour in Demographic Agent-Based Models}
% author commented for blind review purposes
%\author{Jonathan Gray \and Jason Hilton \and Jakub Bijak}

@@

We suggest that inclusion and operationalisation of context are best tackled through an iterative process of development, and evaluation against empirical data, somewhat similar to that proposed by \cite{Cioffi-Revilla2010}. This approach places considerable demands on the modeller in terms of developing the model, but also in exercising their judgement as to what should be included, and identifying where a component should be iterated out. In a sense, this is a similar process to the stepwise approaches commonly used in constructing regression models. Here, however, it is considerably less amenable to automation, both because augmenting a model is time consuming, and because evaluating complex models poses considerable challenges. The challenge of evaluation can be somewhat ameliorated by the use of techniques like sensitivity analysis (see \citet{Thiele2014} for a review of several techniques, and \citet{Oakley2004} for an alternative approach), which can support the modeller in identifying how the individual parameters of a model contribute to overall outputs. This too has limitations, because as we discuss later in relation to heterogeneity in agents (section \ref{heterogeneity-in-decision-strategies}), context can take the form of both process, and parameter.
As a result, the onus remains on the modeller to use their best judgement.

@@

As we have intimated, we believe that introducing such additional complications to a simulation necessitates an iterative approach \citep{Cioffi-Revilla2010}. We also believe that the implemented model of decision making should be treated as a feature to iterate on, and that different models should be contrasted against one another \citep{Rossiter2014}. This suggests that modularity should be a key consideration in model design, to support the modeller in evaluating the relative merits of different decision models, given the purpose of the overall research.

@@

That said, as demographers, we are generally concerned with decisions which affect the life-course. These necessarily have consequences which extend far beyond the present \citep{Willekens2001}. For example, the decision to have a child may affect not only a parent's immediate satisfaction with life, labour market participation, income streams, consumption and so forth, but will continue to do so for the rest of their life. Future `opportunity costs' must also be borne in mind \citep{Butz1979}: taking such a decision now may widen or narrow future options in various other areas of life. Similar considerations are salient for migration and partnership decisions, and indeed for many other demographic and social phenomena, for which the questions of time and sequencing of events tend to be too often ignored (see \citet{Abbott2001} for an excellent overview).

@@

Here, \(U^t\), denoting present utility at time \(t\), is a function of the stream of individual rewards \(u_t\) at each time point, weighted  according to \(\delta^{\tau}\), with \(\delta\) being the discounting factor. This form has nice properties in that it is time-consistent; agents who prefer one option over another today will do so until the reward has been realised. As an example, consider a utility maximising individual who saves money to fund a migration attempt, and must choose between two destinations - one which is closer and thus cheaper to travel to, while the alternative is more distant and thus requires a longer saving period, but is also more attractive. If such a migrant discounts future rewards exponentially, once they have made a choice, such a decision will remain optimal until the trip is made, assuming the material circumstances do not change.

@@

Such behaviour can be approximated by the use of quasi-hyperbolic or present-weighted exponential discounting, which, as the names suggests, accounts for time-inconsistency by including a weight on present rewards in the function, either fixed or variable \citep{Benhabib2010}. To return to the example of the migrant above, a hyperbolically or quasi-hyperbolically discounting agent may begin to save with the more distant, more attractive destination in mind, but actually move to the closer host country once they had saved enough for it to be an immediate option, even without receiving any new information about either.

@@

To give a demographic example of how these models of time-preference could be important, the work of  \cite{Wrede2011} is instructive. He investigated the effect of hyperbolic discounting on fertility in a simple three period model where mothers could have both immediate rewards from motherhood, and delayed rewards in the form of being looked after in old age. He found that the introduction of hyperbolic discounting reduces numbers of births if this second, investment, motive is dominant, but that the ability to offset the immediate costs of motherhood through the financial tools might mitigate these effects. The idea of pre-commitment is also mentioned (although not modelled); in this context, an individual might limit their future freedom of action through sterilisation. Agent based models are flexible enough to embed such time-preference models in their wider social context (discussed in Section \ref{sec:context_matters}).

@@

@@

As well as differences between agents, theories of frame and script selection also suggests difference within single agents over time; difference frames and scripts may be selected in response to different contexts, developing experiences, or influences from peers. This leads to discussion of another form of difference in decision making which may be important to demographers, that of differences over the life course. Migrants who were once intending to save for a period and return home to their family may change their mind and settle permanently in response to a new interpretation of their identity and circumstances \citep{Constant2002}.

@@

All of these problems are far from trivial, particularly as there is little consensus about which models of decision are best in which contexts. How then should we proceed with modelling when the fundamental questions about how individuals make decisions, how they treat time, and how they differ between themselves in these factors are still very uncertain? One way forward is to use the same simulation set-up to analyse multiple models of behaviour \citep{Epstein2013, Rossiter2014}. This multi-model approach allows the modeller to attempt to distinguish between more and less plausible models of behaviour in the particular research context. Conditional on the simulation set-up being appropriate, more plausible models may be better able to reproduce multiple empirical patterns at varying scales \citep{Werker2004, Bianchi2008}.

This commitment to analysing multiple models of behaviour within a single simulation project is facilitated by a modular approach to simulation design, so that different models of behaviour can be swapped in and out without changing the underlying conditions of the simulation \citep{Epstein2013}, as can be seen in \citet{Gray2016}. Such an approach reflects standard object-oriented design principles, but is not always reflected in an academic context (see also \citet{Rossiter2015} for discussion of the application of software engineering principles to social simulation). Modularity also allows for the possibility of easy extension by other researchers with access to the code, allowing for further behavioural models to be examined within the same simulation framework.

@@

Second, the existing methods of the statistical design of experiments (e.g. \citeauthor{Chaloner} \citeyear{Chaloner}) would allow for adopting a systematic approach to model design and data collection. In this way, following the suggestions of \citet{Courgeau}, the simulations would be built iteratively, by applying the experimental design principles to identifying the elements of the model which require additional empirical information, and then by collecting this information.  In this way, the important aspects of the decision process discussed above – context, time and heterogeneity – can be explicitly built into the model, through its specification or parameters. Then, they can be learnt about through repeated collection of information – including through experiments. The modularity of the model would help this process, as it would allow for fine-tuning the different aspects of the model independently, before they can be assembled together. An outline of the process is illustrated in Figure \ref{fig:approach}.
